---
title: "Test STING IDM v0.4"
author: "Nick Isaac"
date: "14/02/2023"
output: html_document
---

In this docuent, we're exploring the properties of the model proposed for Minimum Viable Scheme for EU Pollinator Monitoring Scheme. In this iteration, there are two data types: pan traps and transect walks. Both data types are surveyed on the same day each year. There are `v` visits to `n` sites: both numbers can be modified. We first define the model, then simulate data using user-defined parameters. We then try to recover these parameters for one simulated dataset. At present, the only parameters shared between the data types are those the the detection phenology.

Next steps:
- add a covariate on the state model that is shared between data types.
- parameterize simulations using data from the pilot studies

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load the packages, installing where required
#required.packages <- c("nimble", "ggmcmc", "coda", "reshape2", "ggplot2")
#installedPackages <- installed.packages()
#packages.to.install <- setdiff(required.packages, installedPackages)
#  install.packages(packages.to.install)
#if(length(packages.to.install) > 0)  # can cause a crash
library(nimble)
library(ggmcmc)
library(coda)
library(ggplot2)
library(reshape2)
```

Set some global parameters
```{r SetSimPars}
nSites <- 10
nVisits <- 2
maxDate <- 250 # latest survey date. 250 = mid Sept
minDate <- 100 # earliest survey date. 10 = mid April
sim.pars <- list(sigma.psi = 3, # stdev among sites in cloglog occupancy
              alpha.s = 0, # state intercept
              beta.s = 1, # state covariate
              cov = rnorm(n = nSites),
              alpha.p = 0.8, # logit probability of detection 
              beta1 = 200, # date when detectability is highest
              beta2 = 20, # standard deviation of detection phenology
              phScale = 50) #
```

Define the model code in Nimble
```{r}
obsMod <- FALSE
modelcode <- nimbleCode({
    ######################### state model
  for (i in 1:nsite) {
    linPred[i] <- alpha.s + beta.s * cov[i] #+ eta[i]
    log(lambda[i]) <- linPred[i] # simulation below returns NAs unless I make this a fixed number, e.g. 3
    cloglog(psi[i]) <- linPred[i] #  simulation below returns NAs unless I make this a fixed number, e.g. 3
    z[i] ~ dbern(psi[i]) # True occupancy status
    #eta[i] ~ T(dnorm(mu.lambda, tau.psi), 0, Inf)
    #eta[i]~T(dt(0, 1, 1), 0, Inf) # Half Cauchy 
    #eta[i] ~ dnorm(0, tau.psi)
    #eta[i] ~ dnorm(0, sd = sigma.psi)
  }  
  
  ######################### state model priors
  alpha.s ~ dnorm(0, 0.0001)
  beta.s ~ dnorm(0, 0.0001)
  sigma.psi ~ T(dt(0, 1, 1), 0, Inf) # Half Cauchy

if(obsMod){ # for debugging
  ######################### Obs model
  for (i in 1:nsite) {
    for (j in 1:nvisit) {
      
      ##### pan traps
      y1[i,j] ~ dbin(size = nT[i,j], prob = Py[i,j]) # Observed data
      Py[i,j] <- z[i] * p1[i,j]
      logit(p1[i,j]) <- alpha.p #+ logit(pThin[i,j])
      #p1[i,j] <- alpha.p * pThin[i,j]
      
      ##### transects 
      y2[i,j] ~ dpois(lambdaThin[i,j]) # Observed counts
      lambdaThin[i,j] <- lambda[i] * pThin[i,j]

      #### shared: phenology
      logit(pThin[i,j]) <- phScale * f_JD[JulDate[i,j]]
  }}
}
  ######################### Obs model priors
  phScale ~ dnorm(0, 0.0001)
  alpha.p ~ dnorm(-2, 0.0001) # p1 is detection probability per pan trap at peak phenology; could replace with a logistic model
  
  ######################### Seasonality shared effect
  beta1 ~ dunif(1, 365) # peak detectability/activity
  beta2 ~ T(dt(0, 1, 1), 0, 500) # Half Cauchy. Stdev of phenology. At sd=500 the curve is entirely flat

  for (d in 1:365){
    f_JD[d] <- 1/((2*3.14159265359)^0.5 * beta2) * exp(-((d - (beta1))^2 / (2* beta2^2)))
    # could simplify this and evaluate only for dates in the dataset
  }
  
  #########################  derived parameters
  psi.fs <- mean(z[1:nsite])
})
```


Next step is to simulate some data for that model
```{r}
# set the parameters for a simulation
mcSim <- nimbleModel(modelcode,
                     constants = list(nsite = nSites, 
                                   nvisit = nVisits,
                                   sigma.psi = sim.pars["sigma.psi"],
                                   beta1 = sim.pars["beta1"],
                                   beta2 = sim.pars["beta2"],
                                   alpha.s = sim.pars["alpha.s"],
                                   beta.s = sim.pars["beta.s"],
                                   cov = sim.pars$cov,
                                   phScale = sim.pars["phScale"],
                                   alpha.p = sim.pars["alpha.p"],
                                   nT = array(data = 5, dim = c(nSites, nVisits)),
                                   JulDate = round(array(runif(n = nSites*nVisits, 
                                                               min = minDate, max = maxDate), 
                                                         dim = c(nSites, nVisits)), 0)
                                )
                     #inits = as.list(sim.pars)
                     )

CmcSim <- compileNimble(mcSim) # note that we compile the operational model not the code!


nodesToSim <- CmcSim$getDependencies(c("psi", "lambda"),# "p1", "pThin"),
                                           self = F, downstream = T)
nodesToSim
CmcSim$simulate(nodesToSim)

head(CmcSim$y1)[, 1:2]
head(CmcSim$y2)[, 1:2]

##### temp

#nodesToSim <- CmcSim$getDependencies(c("psi", "lambda", "p1", "pThin"),
#                                           self = F, downstream = T)
nodesToSim
#head(CmcSim$eta)
head(CmcSim$psi)
head(CmcSim$z)
```

I think I finally understand the simulation: I've fixed the values of everything in the simulation except `y1` and `y2`.


```{r exploreModel, include=FALSE}
#CmcSim$isData('y1') # FALSE

# plot the dependencies of the model in a graph (only makes sense when the model is very simple)
#library(igraph)
#plot(mcSim$getGraph())

# Determining the nodes and variables in a model
#CmcSim$getVarNames()

#note that lifted nodes can also correspond to different parametrizations
#CmcSim$getNodeNames()

# Query the model's relationships
#CmcSim$getDependencies(c('psi'))
```


Summarise the simulated data
```{r }
occSites <- with(CmcSim, apply(cbind(y1), 1, max)>0)
(simSum  <- c(
    naiveOcc = mean(occSites),
    reportingRate = mean(with(CmcSim, y1/5)), # per pan trap
    meanCount = mean(CmcSim$y2),
    reportingRate_1 = mean(with(CmcSim, y1/5)[occSites,]), # per pan trap
    meanCount_1 = mean(CmcSim$y2[occSites,])
))
```


Now run the model
```{r runModel}
# step 1 create an operational from from NIMBLE/JAGS/BUGS code
model <- nimbleModel(code = modelcode, 
                     constants = list(nsite = nSites, 
                                      nvisit = nVisits,
                                      nT = array(data = 5, dim = c(nSites, nVisits)),
                                      JulDate = round(array(runif(n = nSites*nVisits, 
                                                                  min = minDate, max = maxDate), 
                                                             dim = c(nSites, nVisits)), 0)
                                      ),
                     data = list(y1 = CmcSim$y1), 
                     inits = list(z = apply(CmcSim$y1 > 0, 1, max), 
                                  mu.psi = simSum["naiveOcc"], 
                                  mu.lambda = simSum["meanCount_1"],
                                  alpha.p = simSum["reportingRate_1"],
                                  beta1 = 180,
                                  beta2 = 50)
                     )

# step 2 build an MCMC object using buildMCMC(). we can add some customization here
occMCMC <- buildMCMC(model, 
                     monitors = c('mu.psi', "mu.lambda", 'alpha.p', "beta1", "beta2"), 
                     thin = 3, 
                     useConjugacy = FALSE) # useConjugacy controls whether conjugate samplers are assigned when possible
#about 5 seconds

# step 3 before compiling the MCMC object we need to compile the model first
Cmodel <- compileNimble(model) # NJBI: I don't understand why this step is necessary
# 25 seconds (less for fewer nodes)

# now the MCMC (project = NIMBLE model already associated with a project)
CoccMCMC <- compileNimble(occMCMC, project = model)
# instantaneous

# and now we can use either $run or runMCMC() on the compiled model object.
system.time(
runMCMC_samples <- runMCMC(CoccMCMC, nburnin = 7500, niter = 10000, nchains = 3, samplesAsCodaMCMC = T)
) # 17 seconds for 10K iterations with 3 visits/site
```

Plot the MCMC traces
```{r diagnoseModel}
plot1 <- runMCMC_samples %>% ggs() %>% ggs_traceplot(greek = TRUE)
plot2 <- runMCMC_samples %>% ggs() %>% ggs_density(greek = TRUE)
gridExtra::grid.arrange(plot1, plot2, ncol=2)
```

With the IDM, the phenology parameters are estimated far better than with the occupancy model.

Compare with simulated values
```{r}
pars <- as.data.frame(summary(runMCMC_samples)$statistics)
pars$simulated <- sim.pars[match(rownames(pars), names(sim.pars))]
pars[,c(ncol(pars),1:(ncol(pars)-1))]
```

It would be quite useful to have the median here as well.

Some other diagnostics
```{r diagnostics}
# Rhat
gelman.diag(runMCMC_samples) # fails if psi.fs is included

# Autocorrelation plots
#autocorr.plot(runMCMC_samples)
```


Plot detection phenology
```{r}
beta1 <- melt(sapply(runMCMC_samples, function(x) x[,"beta1"]))$value
beta2 <- melt(sapply(runMCMC_samples, function(x) x[,"beta2"]))$value

jul_dates <- round(seq(from = 1, to = 365, length.out = 50),0)

pDet <- melt(sapply(jul_dates, function(jd){
    (1/((2*pi)^0.5 * beta2) * exp(-((jd - beta1)^2 / (2* beta2^2))))
   }))
    
pDetSumm <- pDet %>% 
  group_by(Var2) %>%
  summarise(mean = mean(value),
            lowerCI = quantile(value, 0.025),
            upperCI = quantile(value, 0.975)
            ) %>%
  ungroup()

pDetSumm$JulDate <- jul_dates[pDetSumm$Var2]

(detPlot <- ggplot(data = pDetSumm, (aes(x=JulDate))) +
  geom_line(aes(y=mean)) +
  geom_ribbon(aes(ymin=lowerCI, ymax=upperCI), alpha=0.2) +
  theme_bw() +
  xlab("Julian Date (1 = 1st January)") +
  ylab("Detection modifier"))
```